---
title: "STATS 302 - Assignment 3"
author: "Jonathan Kumarich"
date: "18/05/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

### Packages and clear
```{r setup}
rm(list=ls())

pacman::p_load(
  ggplot2,
  rstudioapi,
  tinytex,
  dplyr,
  purrr,
  vegan,
  MASS
)

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("mixOmics")

library(mixOmics)

```

### Data Import
```{r data-import}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()

fungal <- read.csv("fungal-all.csv")
fungal.desc <- fungal[, 1:5]
fungal.val <- fungal[ ,-c(1:5)]
rm(fungal)
```

### Question 1
```{r question-1}
pvals <- rep(NA, ncol(fungal.val))

for (i in 1:ncol(fungal.val)) {
  test <- kruskal.test(fungal.val[, i] ~ fungal.desc$Condition)
  pvals[i] <- test$p.value
}; rm(test)

ggplot(data.frame(pvals), aes(x=pvals)) + 
  geom_histogram(fill="#004775", color="black", bins=20) + 
  theme_classic() +
  xlab("P-Values") + 
  ylab("Count") + 
  ggtitle("Distribution of P-Values") +
  theme(plot.title = element_text(hjust = 0.5)) 

```

If none of the metabolites differed, then by chance, we would have expected around 5% of values to fall under the 0.05 p-value threshold. This would equate to the histogram generally being flat over the whole range of 0 - 1.00. Instead we see a large proportion of p-values appearing in the 0.00 - 0.05 range, indicating that there may be multiple metabolites that differ across conditions.

```{r question-1-qvalues}
qval.pi0<-function(pvals, pi0=1){
  #pvals should not have missing values
  #default value of pi0 is 1, but another can be specified.
  ord<-order(pvals, decreasing = TRUE)
  pord<-pvals[ord]
  answer.ord<-rep(NA,length(pvals) )
  answer<-rep(NA, length(pvals))
  denominator<-sum(pord <= pord[1]) #number of discoveries using pord[i] as the threshold
  numerator<- pord[1]*length(pord)*pi0  #expected number of false discoveries
  answer.ord[1]<-numerator/denominator
  for(i in 1: length(pord)){
    denominator<-sum(pord <= pord[i])
    numerator<- pord[i]*length(pord)*pi0
    if( numerator/denominator <= min(answer.ord[1:(i-1)]) ) answer.ord[i]<-numerator/denominator
    if( numerator/denominator > min(answer.ord[1:(i-1)]) ) answer.ord[i]<- min(answer.ord[1:(i-1)])
  }
  answer[ord]<-answer.ord
  return(answer)
}

qvals <- qval.pi0(pvals)

qvals.significant <- sum(qvals < 0.05)

print("Number of observations with significant q-values: " %>% paste0(qvals.significant))


```

```{r question-1-q&pvals}

qandp.significant <- sum((qvals >= 0.05) * (pvals < 0.05))

print("Number of observations with significant p-values but non-significant q-values: " %>% paste0(qandp.significant))

```

### Question 2

```{r question-2}

plsda.output <- plsda(X=fungal.val ,Y=fungal.desc$genera, scale=TRUE, ncomp=11) 

plotVar(plsda.output, cex=2)

```

There appears to be a large number of variables that are very highly associated with Component 1 (with a slight negative association to Component 2), as well as a large number of variables that have a very high association to Component 1 (with a slight positive association to Component 1. There also appears to be a significant number of variables that have very low associations with both components.

```{r question-2-plotindiv}

plotIndiv(plsda.output, comp=1:2, ind.names=F, legend=T)

```

We can see from the plot of individuals that a lot of the genera types have been separated out. Genera such as Kluyveromyces, Clavispora, and Trichosporon, all have very high levels of X-variate 1, but all are quite hard to distinguish from each other within this sub-group (except Clavispora which has a lower X-variate 2). On the other hand, the Broth, Scedosporium, and Saccharomyces all have relatively low levels of X-variate 1. Exophiala and Coniochaeta, both have X-variate 1 values centered around zero.

Candida and Pichia are quite hard to distinguish as they seem relatively unrelated to both X-variate 1 and 2.

### Question 3
```{r question-3} 

plsda.perf <- plsda.output %>% perf(validaton='loo', folds=6, nrepeat=10)

plsda.perf %>% plot()

```

```{r question-3-BER}

plsda.perf$error.rate$BER

```

We can see that the balanced error rate is the lowers for the Mahalanobis distance at 11 components, indication that we should use all 11 components.

We want to use the balanced error rate from the plot as the groups are of different sizes and we would like them to all have equal weight when determining the error rate.

### Question 4
```{r question-4}

funsub <- read.csv("funsub-assignment.csv")

condition <- funsub$Condition
metabolites <- funsub[, 5:ncol(funsub)]

mod <- manova(as.matrix(metabolites) ~ condition)

W <- summary(mod)$SS$Residual

Sigma <- data.frame(W / (nrow(metabolites) - 2))
```


```{r question4-group-centers}

apply(funsub[funsub$Condition=="d_SCFA",5:28],2,mean)->SCFA.mean
apply(funsub[funsub$Condition=="Control",5:28],2,mean)->Ctrl.mean
matmean<-matrix(rep(NA,24*108), ncol=24)
for(i in 1:108){
if(funsub$Condition[i]=='d_SCFA'){  
matmean[i,]<-SCFA.mean}
  else{
matmean[i,]<-Ctrl.mean}}
funsub.center<-funsub[,5:28]-matmean

distances <- mahalanobis(funsub.center, 0, Sigma) 
hist(distances)

```

```{r question-4-cov-test}

kruskal.test(as.matrix(funsub.center)[,2] ~ funsub$Condition)

pvals <- rep(NA, ncol(funsub.center))

for (i in 1:ncol(funsub.center)) {
  test <- kruskal.test(funsub.center[, i] ~ funsub$Condition)
  pvals[i] <- test$p.value
}; rm(test)

ggplot(data.frame(pvals), aes(x=pvals)) + 
  geom_histogram(fill="#004775", color="black", bins=20) + 
  theme_classic() +
  xlab("P-Values") + 
  xlim(0,1)
  ylab("Count") + 
  ggtitle("Distribution of P-Values") +
  theme(plot.title = element_text(hjust = 0.5)) 
```

Looking at the VanValens test for all of the metabolites, we can see that all of them have non-significant p-values which indicate that the within group covariances are similar.


```{r questino-4-qqplot}

qqplot(
  qchisq(
    ppoints(nrow(funsub.center)), 
    df=ncol(funsub.center)), 
  distances, 
  main="Mahlanobis distances vs Chi-squared distribuiton", 
  xlab='Chisq Quantiles') 
abline(0,1)

```

The quantiles of distances appear to follow a chi-squared distribution which also provides evidence for multivariate normality.


```{r question-4-dist-test}

ks.test(distances, "pchisq", df=ncol(funsub.center))

```

Here we have a non-significant p-value (0.381) which further indicates multivariate normality.


### Question 5

```{r question-5}

summary(mod, test="Pillai", df = ncol(funsub))


```

Using the MANOVA fit earlier, we calculated Pillai's trace of the metabolites with respect to Condition. We got a highly significant p-value indicating a strong relationship between the condition and the metabolite values. This fits the assumptions of the MANOVA, since there was similar within-group covariances and multivariate normality, as seen previously in question 4. 



